Discrete Quantum Theories and Computing

### 2018-11-20 dissertation.tex
dissertation.lyx "2.2 Quantum Probability" Updating...

A \emph{probability space} is a mathematical abstraction specifying
the necessary conditions for reasoning coherently about collections
of uncertain events \cite{Kolmogorov1950,544199,Griffiths2003,Grabisch2016}.
Given a finite sample space~$\Omega$ representing all possible outcomes
of a process, and its power set $2^{\Omega}$ as the classical event
space, a classical probability measure~$\mu$ maps an event~$E$
to a number $\mu\left(E\right)\in\left[0,1\right]$ specifying how
likely one of the outcome in $E$ actually happens. To maintain the
coherence, $\mu$ is subject to the following constraints: $\mu\left(\emptyset\right)=0$,
$\mu\left(\Omega\right)=1$, $\mu\left(\overline{E}\right)=1-\mu\left(E\right)$,
and $\mu\left(E_{0}\cup E_{1}\right)=\mu\left(E_{0}\right)+\mu\left(E_{1}\right)$
for each pair of disjoint events $E_{0}\subseteq\Omega$ and $E_{1}\subseteq\Omega$.

Since the previous abstraction doesn't specify how to generate the
outcomes, the generating process could well be a quantum experiment.
Let us prepare a beam of one kind of spin~$1$ particles with basis
vectors $\ket{0}$, $\ket{1}$, and $\ket{2}$. In principle, this
beam can be split by a Stern-Gerlach type experiment according to
the eigenvalues of the observable $\mathbf{O}_{0}=0\,\proj{0}+1\,\proj{1}+2\,\proj{2}$,
and the states corresponding to the split beam are $\ket{0}$, $\ket{1}$,
and $\ket{2}$ \cite{:/content/aip/journal/jmp/21/1/10.1063/1.524312,peres1995quantum}.
Instead of a beam, if only one particle is sent to the beam splitter,
the state after the experiment is one of $\ket{0}$, $\ket{1}$, and
$\ket{2}$, and each state has a probability. In the language of our
previous abstraction, the set of outcomes is $\Omega_{0}=\left\{ \ket{0},\ket{1},\ket{2}\right\} $.
All possible events are $\emptyset$, $\left\{ \ket{0}\right\} $,
$\left\{ \ket{1}\right\} $, $\left\{ \ket{2}\right\} $, $\left\{ \ket{0},\ket{1}\right\} $,
$\left\{ \ket{0},\ket{2}\right\} $, $\left\{ \ket{1},\ket{2}\right\} $,
and $\Omega_{0}$. This experiment defines a classical probability
measure $\mu_{0}:2^{\Omega_{0}}\rightarrow\left[0,1\right]$.

One special feature of quantum experiments is that the probabilities
in different experiments are correlated. Consider another experiment
similar to the previous one. While this experiment send exactly the
same particle, we use a beam splitter corresponding to the observable
$\mathbf{O}_{1}=0\,\proj{\ps}+1\,\proj{\ms}+2\,\proj{2}$, where $\ket{\ps}=\frac{\ket{0}+\ket{1}}{\sqrt{2}}$
and $\ket{\ms}=\frac{\ket{0}-\ket{1}}{\sqrt{2}}$. The outcome of
this experiment is $\Omega_{1}=\left\{ \ket{\ps},\ket{\ms},\ket{2}\right\} $,
and it defines another probability measure $\mu_{1}:2^{\Omega_{1}}\rightarrow\left[0,1\right]$.
Although $\mu_{0}$ and $\mu_{1}$ are different probability measures,
the probability of the same event are experimentally found to be the
same, i.e., $\mu_{0}\left(\emptyset\right)=\mu_{1}\left(\emptyset\right)$
and $\mu_{0}\left(\left\{ \ket{2}\right\} \right)=\mu_{1}\left(\left\{ \ket{2}\right\} \right)$.
In general, when the particle and the event are the same, so does
their probability no matter how we set up the beam splitters.

Since the probability induced by the different beam splitters are
correlated, it is more natural to define the set of quantum event~$\events$
as the union of all possible event spaces using different beam splitters.
Since we are focus on the probability instead of outcome, and the
probabilities of events $\left\{ \ket{0},\ket{1}\right\} $ and $\left\{ \ket{\ps},\ket{\ms}\right\} $
are always the same, it is reasonable to identify them as the same
event. To simplify the notation, given any classical event $E\subseteq\Omega$,
its corresponding quantum event is defined by 
\begin{equation}
\varphi\left(E\right)=\sum_{\ket{j}\in E}\proj{j}\,.\label{eq:pullback-function}
\end{equation}
Following this definition, it is easy to check that not only 
\begin{equation}
\varphi\left(\left\{ \ket{0},\ket{1}\right\} \right)=\proj{0}+\proj{1}=\proj{\ps}+\proj{\ms}=\varphi\left(\left\{ \ket{\ps},\ket{\ms}\right\} \right)
\end{equation}
but every classical event mapping the same projector must also have
the same probability. If we follow the convention $\varphi\left(\emptyset\right)=\mathbb{0}$,
this function $\varphi$ naturally sends the set operations to the
corresponding projector operations: $\varphi\left(\Omega\right)=\mathbb{1}$,
$\varphi\left(\overline{E}\right)=\mathbb{1}-\varphi\left(E\right)$.
For each pair of set $E_{0}\subseteq\Omega$ and $E_{1}\subseteq\Omega$,
\begin{align}
\varphi\left(E_{0}\cap E_{1}\right) & =\varphi\left(E_{0}\right)\varphi\left(E_{1}\right)\,, & \varphi\left(E_{0}\cup E_{1}\right) & =\varphi\left(E_{0}\right)+\varphi\left(E_{1}\right)-\varphi\left(E_{0}\right)\varphi\left(E_{1}\right)\,.\label{eq:pullback-properties}
\end{align}
If $E_{0}$ and $E_{1}$ are disjoint, then $\varphi\left(E_{0}\right)$
and $\varphi\left(E_{1}\right)$ are orthogonal, where two projectors
$P_{0}$ and $P_{1}$ are called \emph{orthogonal} if $P_{0}P_{1}=\mathbb{0}$.

, and consider the quantum probability measure $\mu:\events\rightarrow\left[0,1\right]$
as a function assigning the likelihood to every quantum event

In the quantum case, the events of interest are specified by \emph{projection
operators} $P$ satisfying the condition $P^{2}=P$. These include
the empty projector~$\mathbb{0}$, the identity projector~$\mathbb{1}$,
projectors of the form $\proj{\Phi}$ where $\ket{\Phi}$ is a pure
quantum state (an element of a Hilbert space $\mathcal{H}$), sums
of \emph{orthogonal} projectors~$P_{0}$ and~$P_{1}$ with $P_{0}P_{1}=\mathbb{0}$,
and products of \emph{commuting} projectors $P_{0}$ and~$P_{1}$
with $P_{0}P_{1}=P_{1}P_{0}$. In a quantum probability space \cite{10.2307/2308516,gleason1957,Redhead1987-REDINA,Maassen2010,Abramsky2012},
the set of all events is the set of all projectors on a given Hilbert
space denoted by $\events$. \todo{Use the same idea as the classical
probability measure to explain the meaning of $\mu\left(P\right)$.
Maybe move the structure preserving function $\varphi:2^{\Omega}\rightarrow\events$
defined in Eq.~\ref{eq:pullback-function} here.} each event~$P_{i}$
is mapped to a probability~$\mu\left(P_{i}\right)$ using a probability
measure $\mu:\events\rightarrow\left[0,1\right]$ subject to the following
constraints:
\begin{align}
\mu\left(\mathbb{0}\right) & =0\,, & \mu\left(\mathbb{1}\right) & =1\,, & \mu\left(\mathbb{1}-P\right) & =1-\mu\left(P\right)\,,\label{eq:QuantumProbability}
\end{align}
and for each pair of \emph{orthogonal} projectors $P_{0}$ and $P_{1}$:
\begin{equation}
\mu\left(P_{0}+P_{1}\right)=\mu\left(P_{0}\right)+\mu\left(P_{1}\right)\,.\label{eq:QuantumProbability-Addition}
\end{equation}
Given a Hilbert space $\mathcal{H}$ of dimension $D$ and a probability
assignment for every projector $P$, we can define the expectation
value of an observable~$\mathbf{O}$ having spectral decomposition
$\mathbf{O}=\sum_{i=0}^{D-1}\lambda_{i}P_{i}$, with eigenvalues $\lambda_{i}\in\mathbb{R}$,
as \cite{544199,Jaeger2007}: 
\begin{equation}
\expval{\mathbf{O}}_{\mu}=\sum_{i=0}^{D-1}\lambda_{i}\mu\left(P_{i}\right)\,,\label{eq:quantum-expectation}
\end{equation}
where the subscript $\mu$ might be omitted if it is clear according
to the context.

\todo{Rewrite the beginng of this paragraph to explain the meaning
of the Born rule... Maybe move Gleason's Theorem here?} A conventional
quantum probability measure can easily be constructed using the Born
rule if one knows the current pure unnormalized quantum state $\ket{\Phi}\in\mathcal{H}$;
then the Born rule induces a probability measure $\muB_{\Phi}$ defined
as 
\begin{equation}
\muB_{\Phi}(P)=\frac{\melem{\Phi}{P}{\Phi}}{\ip{\Phi}{\Phi}}\,.\label{eq:Born}
\end{equation}
If $\ket{\Phi}$ is normalized, Eq.~(\ref{eq:Born}) could be simplified
as $\muB_{\Phi}(P)=\melem{\Phi}{P}{\Phi}$. For mixed states $\rho=\sum_{j=1}^{N}q_{j}\proj{\Phi_{j}}$,
where $\ket{\Phi_{j}}\in\mathcal{H}$ are normalized, $q_{j}>0$,
and $\sum_{j=1}^{N}q_{j}=1$, the generalized Born rule induces a
probability measure $\muB_{\rho}$ defined as \cite{Born1983,Mermin2007,Jaeger2007}
\begin{equation}
\muB_{\rho}\left(P\right)=\Tr\left(\rho P\right)=\sum_{j=1}^{N}q_{j}\muB_{\Phi_{j}}\left(P\right)\,.\label{eq:mixed-Born}
\end{equation}

>>>>

A \emph{probability space} is a mathematical abstraction specifying
the necessary conditions for reasoning coherently about collections
of uncertain events \cite{Kolmogorov1950,544199,Griffiths2003,Grabisch2016}.
Given a finite sample space~$\Omega$ representing all possible outcomes
of a process, and its power set $2^{\Omega}$ as the classical event
space, a classical probability measure~$\mu$ maps an event~$E$
to a number $\mu\left(E\right)\in\left[0,1\right]$ specifying how
likely one of the outcome in $E$ actually happens. To maintain the
coherence, $\mu$ is subject to the following constraints: $\mu\left(\emptyset\right)=0$,
$\mu\left(\Omega\right)=1$, and $\mu\left(\overline{E}\right)=1-\mu\left(E\right)$,
where $\overline{E}$ is the complement of $E$.  Moreover, $\mu\left(E_{0}\cup E_{1}\right)=\mu\left(E_{0}\right)+\mu\left(E_{1}\right)$
for each pair of disjoint events $E_{0}\subseteq\Omega$ and $E_{1}\subseteq\Omega$.

Since the previous abstraction doesn't specify the process to generate
the outcomes, this process could well be a quantum experiment. Let
us prepare a beam of one kind of spin~$1$ particles with basis vectors
$\ket{0}$, $\ket{1}$, and $\ket{2}$. In principle, this beam can
be split by a Stern-Gerlach type experiment according to the eigenvalues
of the observable $\mathbf{O}_{0}=0\,\proj{0}+1\,\proj{1}+2\,\proj{2}$,
and the states corresponding to the split beams are $\ket{0}$, $\ket{1}$,
and $\ket{2}$ \cite{:/content/aip/journal/jmp/21/1/10.1063/1.524312,peres1995quantum}.
Instead of sending a beam, if only one particle is sent to the beam
splitter, the state after the experiment is one of $\ket{0}$, $\ket{1}$,
and $\ket{2}$, and there is a probability for each post-experimental
state. In the language of our abstraction, the set of outcomes is
$\Omega_{0}=\left\{ \ket{0},\ket{1},\ket{2}\right\} $. All possible
events are $\emptyset$, $\left\{ \ket{0}\right\} $, $\left\{ \ket{1}\right\} $,
$\left\{ \ket{2}\right\} $, $\left\{ \ket{0},\ket{1}\right\} $,
$\left\{ \ket{0},\ket{2}\right\} $, $\left\{ \ket{1},\ket{2}\right\} $,
and $\Omega_{0}$. And this experiment defines a classical probability
measure $\mu_{0}:2^{\Omega_{0}}\rightarrow\left[0,1\right]$.

One special feature of quantum experiments is that the probabilities
in different experiments are correlated. Consider another experiment
similar to the previous one. While sending exactly the same particle,
we use a different beam splitter corresponding to the observable $\mathbf{O}_{1}=0\,\proj{\ps}+1\,\proj{\ms}+2\,\proj{2}$,
where $\ket{\ps}=\frac{\ket{0}+\ket{1}}{\sqrt{2}}$ and $\ket{\ms}=\frac{\ket{0}-\ket{1}}{\sqrt{2}}$.
Now the sample space is $\Omega_{1}=\left\{ \ket{\ps},\ket{\ms},\ket{2}\right\} $,
and another probability measure $\mu_{1}:2^{\Omega_{1}}\rightarrow\left[0,1\right]$
is defined. Although the sample spaces and the probability measures
are different, these two experiments may produce the same outcome
state. Moreover, the probability of the same event in different experiments
are experimentally found to be the same, i.e., $\mu_{0}\left(\emptyset\right)=\mu_{1}\left(\emptyset\right)$
and $\mu_{0}\left(\left\{ \ket{2}\right\} \right)=\mu_{1}\left(\left\{ \ket{2}\right\} \right)$.
In general, as long as sending the same particle, the probability
of the same event in different experiments are always the same. This
fact is equivalent to the fact that commuting observables could be
measured simultaneous, and the latter fact is essential to define
contextuality which will be explained later.

Since the probability induced by the different beam splitters are
correlated, it is more natural to define the set of quantum events~$\events$
as the union of all possible event spaces using different beam splitters
on a given Hilbert space. Because we are focus on the probability
instead of outcome, and the probabilities of events $\left\{ \ket{0},\ket{1}\right\} $
and $\left\{ \ket{\ps},\ket{\ms}\right\} $ are always the same, it
is reasonable to identify them as the same event. By observing the
projectors generated by these two events are the same
\begin{equation}
\proj{0}+\proj{1}=\proj{\ps}+\proj{\ms}\,,
\end{equation}
we define the projector generated by any classical event~$E$ as
its quantum event 
\begin{equation}
\varphi\left(E\right)=\sum_{\ket{j}\in E}\proj{j}\in\events\,.\label{eq:pullback-function}
\end{equation}
If we follow the convention $\varphi\left(\emptyset\right)=\mathbb{0}$,
not only the same projector must have the same probability, but this
function $\varphi$ also naturally sends the set structure to the
corresponding projector structure: $\varphi\left(\Omega\right)=\mathbb{1}$
and $\varphi\left(\overline{E}\right)=\mathbb{1}-\varphi\left(E\right)$.
Given a sample space $\Omega$ and a pair of event $E_{0}\subseteq\Omega$
and $E_{1}\subseteq\Omega$, their corresponding quantum events $\varphi\left(E_{0}\right)$
and $\varphi\left(E_{1}\right)$ are commuting and vice versa, where
two projectors $P_{0}$ and~$P_{1}$ are \emph{commuting} if $P_{0}P_{1}=P_{1}P_{0}$.
Moreover, $E_{0}$ and $E_{1}$ also satisfy the following properties:
\begin{align}
\varphi\left(E_{0}\cap E_{1}\right) & =\varphi\left(E_{0}\right)\varphi\left(E_{1}\right)\,, & \varphi\left(E_{0}\cup E_{1}\right) & =\varphi\left(E_{0}\right)+\varphi\left(E_{1}\right)-\varphi\left(E_{0}\right)\varphi\left(E_{1}\right)\,.\label{eq:pullback-properties}
\end{align}
If $E_{0}$ and $E_{1}$ are disjoint, then $\varphi\left(E_{0}\right)$
and $\varphi\left(E_{1}\right)$ are orthogonal and vice versa, where
two projectors $P_{0}$ and $P_{1}$ are called \emph{orthogonal}
if $P_{0}P_{1}=\mathbb{0}$.

After we define the set of all events~$\events$ as the set of all
projectors on a given Hilbert space, a quantum probability space can
be defined by $\events$ together with a probability measure $\mu:\events\rightarrow\left[0,1\right]$
subject to the following constraints \cite{10.2307/2308516,gleason1957,Redhead1987-REDINA,Maassen2010,Abramsky2012}:
\begin{align}
\mu\left(\mathbb{0}\right) & =0\,, & \mu\left(\mathbb{1}\right) & =1\,, & \mu\left(\mathbb{1}-P\right) & =1-\mu\left(P\right)\,,\label{eq:QuantumProbability}
\end{align}
and for each pair of \emph{orthogonal} projectors $P_{0}$ and $P_{1}$:
\begin{equation}
\mu\left(P_{0}+P_{1}\right)=\mu\left(P_{0}\right)+\mu\left(P_{1}\right)\,.\label{eq:QuantumProbability-Addition}
\end{equation}
Given a Hilbert space $\mathcal{H}$ of dimension $D$ and a probability
assignment for every projector $P$, we can define the expectation
value of an observable~$\mathbf{O}$ having spectral decomposition
$\mathbf{O}=\sum_{i=0}^{D-1}\lambda_{i}P_{i}$, with eigenvalues $\lambda_{i}\in\mathbb{R}$,
as \cite{544199,Jaeger2007}: 
\begin{equation}
\expval{\mathbf{O}}_{\mu}=\sum_{i=0}^{D-1}\lambda_{i}\mu\left(P_{i}\right)\,,\label{eq:quantum-expectation}
\end{equation}
where the subscript $\mu$ might be omitted if it is clear according
to the context.

\todo{Rewrite the beginning of this paragraph to explain the meaning
of the Born rule... Maybe move Gleason's Theorem here?} A conventional
quantum probability measure can easily be constructed using the Born
rule if one knows the current pure unnormalized quantum state $\ket{\Phi}\in\mathcal{H}$;
then the Born rule induces a probability measure $\muB_{\Phi}$ defined
as 
\begin{equation}
\muB_{\Phi}(P)=\frac{\melem{\Phi}{P}{\Phi}}{\ip{\Phi}{\Phi}}\,.\label{eq:Born}
\end{equation}
If $\ket{\Phi}$ is normalized, Eq.~(\ref{eq:Born}) could be simplified
as $\muB_{\Phi}(P)=\melem{\Phi}{P}{\Phi}$. For mixed states $\rho=\sum_{j=1}^{N}q_{j}\proj{\Phi_{j}}$,
where $\ket{\Phi_{j}}\in\mathcal{H}$ are normalized, $q_{j}>0$,
and $\sum_{j=1}^{N}q_{j}=1$, the generalized Born rule induces a
probability measure $\muB_{\rho}$ defined as \cite{Born1983,Mermin2007,Jaeger2007}
\begin{equation}
\muB_{\rho}\left(P\right)=\Tr\left(\rho P\right)=\sum_{j=1}^{N}q_{j}\muB_{\Phi_{j}}\left(P\right)\,.\label{eq:mixed-Born}
\end{equation}
